UKB Models

### 2023-10-11 proof-of-concept_mods.Rmd
fixed summary parsing, visualizing model outputs

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(gamlss, gamlss.cens, gamlss.dist, gamlss.mx, gamlss.tr, ggplot2, tidyverse, ggpubr, skimr, gamlss.data, data.table, nlme, devtools, slider)

@@

```{r}
ukb_df_totals <- read_csv(file="data/ukb_CN_data.csv")
ukb_df_age.filt <- read_csv(file="data/ukb_CN_data_agefilt.csv")
```

@@

```{r}
basic.m1 <- gamlss(formula = GMV ~ poly(age_days, 3) + sex,
                       sigma.formula = ~ poly(age_days, 3) + sex,
                       nu.formula = ~ 1,
                       family = BCCG, data=ukb_df_age.filt, trace = FALSE)
```

@@

```{r}
drop1(basic.m1, what="sigma", scope = "sex")
```

>>>>

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(gamlss, gamlss.cens, gamlss.dist, gamlss.mx, gamlss.tr, ggplot2, tidyverse, ggpubr, skimr, gamlss.data, data.table, nlme, devtools, slider, tidymodels, broom.mixed, ggseg)

@@

```{r}
ukb_df_totals <- read_csv(file="data/ukb_CN_data.csv")
ukb_df_age.filt <- read_csv(file="data/ukb_CN_data_agefilt.csv")

vol_list_global <- readRDS(file="R_scripts/vol_list_global.rds")
vol_list_regions <- readRDS(file="R_scripts/Vol_list_regions.rds")
sa_list <- readRDS(file="R_scripts/SA_list.rds")
ct_list <- readRDS(file="R_scripts/CT_list.rds")
```

@@

```{r}
basic.m1 <- gamlss(formula = GMV ~ poly(age_days, 3) + sex,
                       sigma.formula = ~ poly(age_days, 3) + sex,
                       nu.formula = ~ 1,
                       family = BCCG, data=ukb_df_age.filt, trace = FALSE)
summary(basic.m1)
```

@@

```{r}
K <- drop1(basic.m1, what="sigma", scope = "sex")
class(K)
K["Pr(Chi)"]
K$"Pr(Chi)"[2]

test.mu <- drop1(basic.m1, what="mu", scope = "sex")
test.sig <- drop1(basic.m1, what="sigma", scope = "sex")

#write for testing
saveRDS(basic.m1, file="/Users/megardn/Desktop/BGD_Repos/combat_biovar/gamlss_basic_test_mod.rds")

get.and.drop1.p("/Users/megardn/Desktop/BGD_Repos/combat_biovar/gamlss_basic_test_mod.rds", moment="sigma", term="sex")

test.model.list <- c()
test.model.list[[1]] <- basic.m1
lapply(test.model.list, drop1, what="mu", scope = "sex")
```

## Significance Testing

scp'd models from cubic to parse

```{r}
read_path <- "/Users/megardn/Desktop/BGD_Repos/combat_biovar/ukb_basic/gamlss_objs" #path to bfp models
```

```{r}
#iterate across gamlss objects
model.files <- list.files(path = read_path, pattern = "mod.rds", full.names = TRUE)

sigma.df <- data.frame("file_path" = model.files)
sigma.df <- sigma.df %>%
  mutate(region = sub("_mod\\.rds$", "", basename(as.character(file_path))))

#test significance of sex term in sigma
sigma.df$sig.sex_drop1.p <- lapply(sigma.df$file_path, get.and.drop1.p, moment="sigma", term="sex")

#extract values of sex term from summary table
summary.list <- lapply(sigma.df$file_path, get.summary)
summary.df <- bind_rows(summary.list)
sigma.sex.df <- summary.df %>%
  dplyr::filter(parameter == "sigma" & term == "sexMale") %>%
  mutate(pheno_cat = as.factor(case_when(
    pheno %in% vol_list_global ~ "Global Volume",
    pheno %in% vol_list_regions ~ "Regional Volume",
    pheno %in% sa_list ~ "Regional SA",
    pheno %in% ct_list ~ "Regional CT",
    TRUE ~ NA)),
    sig = case_when(p.value < 0.05 ~ TRUE,
                    p.value >= 0.05 ~ FALSE),
    sig_bf.corr = case_when(pheno %in% vol_list_global & p.value < (0.05/length(vol_list_global)) ~ TRUE,
                            pheno %in% vol_list_global & p.value >= (0.05/length(vol_list_global)) ~ FALSE,
                            !(pheno %in% vol_list_global) & p.value < (0.05/length(ct_list)) ~ TRUE,
                            !(pheno %in% vol_list_global) & p.value >= (0.05/length(ct_list)) ~ FALSE,
                            TRUE ~ NA),
    label = sub("_[^_]*_", "_", pheno)) #for plotting

#add more info from parcellations
dk.parc <- dk$data %>%
  as.data.frame() %>%
  na.omit() %>%
  dplyr::select(c(hemi, region, label)) %>%
  distinct()

sigma.sex.df2 <- left_join(sigma.sex.df, dk.parc, by="label")
```

Visualize

```{r}
table(sigma.sex.df2$sig)
table(sigma.sex.df2$sig_bf.corr)
```

```{r, fig.width=11}
sigma.sex.df2 %>%
  dplyr::filter(sig_bf.corr == TRUE) %>%
  ggplot() +
  geom_col(aes(x=reorder(pheno,t_stat), y=t_stat, fill=pheno_cat)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Significant effects of male sex on sigma across phenotype-specific brain charts")

sigma.sex.df2 %>%
  dplyr::filter(sig_bf.corr == TRUE) %>%
  ggplot() +
  geom_col(aes(x=label, y=t_stat, fill=pheno_cat)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Significant effects of male sex on sigma across phenotype-specific brain charts")

sigma.sex.df2 %>%
  dplyr::filter(sig_bf.corr == TRUE) %>%
  ggplot() +
  geom_col(aes(x=region, y=t_stat, fill=hemi)) + 
  facet_wrap(~pheno_cat) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Significant effects of male sex on sigma across phenotype-specific brain charts")
```

Plot on brains!

```{r}
sigma.sex.df2 %>%
  dplyr::filter(sig_bf.corr == TRUE) %>%
  group_by(pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = t_stat)) +
  scale_fill_gradient2() +
  facet_wrap(~pheno_cat) +
  theme_void() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  ggtitle("t-statistic for effect of male sex on sigma")
```

```{r}

```

### 2023-10-11 ukb_curation-eda.Rmd
fixed summary parsing, visualizing model outputs

#split by measure type (Vol, SA, CT)
vol_list_global <- c("GMV", "sGMV", "WMV", "Ventricles")
writeLines(vol_list_global, con = "pheno_lists/Vol_list_global.txt")

vol_list_regions <- col_list[grepl("^lh_Vol|^rh_Vol", col_list)]
writeLines(vol_list_regions, con = "pheno_lists/Vol_list_regions.txt")

sa_list <- col_list[grepl("^lh_SA|^rh_SA", col_list)]
writeLines(sa_list, con = "pheno_lists/SA_list.txt")

ct_list <- col_list[grepl("^lh_CT|^rh_CT", col_list)]
writeLines(ct_list, con = "pheno_lists/CT_list.txt")
```

>>>>

#split by measure type (Vol, SA, CT)
vol_list_global <- c("GMV", "sGMV", "WMV", "Ventricles")
writeLines(vol_list_global, con = "pheno_lists/Vol_list_global.txt")
saveRDS(vol_list_global, file="R_scripts/vol_list_global.rds")

vol_list_regions <- col_list[grepl("^lh_Vol|^rh_Vol", col_list)]
writeLines(vol_list_regions, con = "pheno_lists/Vol_list_regions.txt")
saveRDS(vol_list_regions, file="R_scripts/Vol_list_regions.rds")

sa_list <- col_list[grepl("^lh_SA|^rh_SA", col_list)]
writeLines(sa_list, con = "pheno_lists/SA_list.txt")
saveRDS(sa_list, file="R_scripts/SA_list.rds")

ct_list <- col_list[grepl("^lh_CT|^rh_CT", col_list)]
writeLines(ct_list, con = "pheno_lists/CT_list.txt")
saveRDS(ct_list, file="R_scripts/CT_list.rds")
```
